ML Estimation of Matrix t Distributions with Expectation Maximization

### 2016-12-11 paper.tex
Make everything an equation

We can split the integral in Eq.~\ref{eq:expdef} by the terms of
$\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})$. Eliminating
a global factor $\frac{1}{2}$ and terms that do not depend on the
maximization parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$,
we have: 
\[
\begin{split}\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\sim n_{t}\log\det S+\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))\\
-n_{t}p\log2-p\log\det\Omega_{t}+(n_{t}+p-1)\log\det\Sigma_{t}-2\log\Gamma_{p}(\frac{1}{2}(n_{t}+p-1))
\end{split}
\]
where the first two terms depend on $S$, and therefore we need to
compute the corresponding integral, while for the rest, the integral
is proportional to the PDF of the matrix--$t$ distribution, Eq.~\ref{eq:matrixtpdf}.
In order to compute:
\begin{equation}
n_{t}\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\log\det SdS\label{eq:firstlogpart}
\end{equation}
we note that taking the derivative with respect to $x$ in Eq.~\ref{eq:intmultgammadef},
we have:
\begin{equation}
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}\log\det SdS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}(\log\det A+\gamma_{p}(\frac{1}{2}(1+p+2x)))\label{eq:intderivative}
\end{equation}
We can solve Eq.~\ref{eq:firstlogpart} by substituting this result directly.

In order to compute the second nontrivial log term,
\[
\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))dS
\]
we note that, due to the linearity of the trace, the result is equivalent
to solving a matrix valued integral
\[
\textrm{tr}\left[\left(\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)SdS\right)(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\,.
\]
The enclosed integral also follows from Eq.~\ref{eq:intmultgammadef},
after taking the matrix derivative with respect to $A$. Using, for
symmetric $S$
\begin{equation}
\frac{d}{dA}\det A=(\det A)A^{-1}\label{eq:detderivative}
\end{equation}
and
\[
\frac{d}{dA}e^{\textrm{tr}SA}=(e^{\textrm{trS}A})S
\]
we have 
\[
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}SdS=-\frac{1}{2}(1+p+2x)\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(p+2x-1)}A^{-1}
\]
from where again, we can obtain the result by direct substitution.

@@

For $M_{t}$ we have:
\[
\frac{dE}{dM_{t}}=0\Rightarrow\sum_{i}^{N}\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}\left(T_{i}-M_{t}\right)\Omega_{t}^{-1}=0
\]
and because $\Omega_{t}$ is nonsingular, we can obtain $M_{t}$ as
a weighted average of the input data:
\[
M_{t}=\frac{1}{N}\sum_{i}^{N}\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}T_{i}=\frac{1}{N}\sum_{i}^{N}W_{i}T_{i}
\]
where we have defined the weight matrices $W_{i}$ as
\[
W_{i}=\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}
\]

$\Sigma_{t}$ is given in terms of the inverse of the mean weight
by:
\[
\Sigma_{t}=\frac{n_{t}+p-1}{n+m+p-1}\left[\frac{1}{N}\sum_{i}^{N}W_{i}\right]^{-1}
\]
Note that it depends on $n_{t}$.

For $\Omega_{t}$ we obtain:
\[
\Omega_{t}=\frac{n+p+m-1}{p}\frac{1}{N}\sum_{i}^{N}(T_{i}-M_{t})W_{i}(T_{i}-M_{t})
\]
and for $n_{t}$, in the case where $\Sigma_{t}$ is a free parameter
that needs to be estimated, we have:
\[
-p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log\det W_{i}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\]
otherwise if $\Sigma_{t}=\Sigma$ is fixed and does not depend on
$n_{t}$:
\[
\log\det\Sigma_{t}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\]

>>>>

We can split the integral in Eq.~\ref{eq:expdef} by the terms of
$\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})$. Eliminating
a global factor $\frac{1}{2}$ and terms that do not depend on the
maximization parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$,
we have: 
\begin{equation}
\begin{split}\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\sim n_{t}\log\det S+\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))\\
-n_{t}p\log2-p\log\det\Omega_{t}+(n_{t}+p-1)\log\det\Sigma_{t}-2\log\Gamma_{p}(\frac{1}{2}(n_{t}+p-1))
\end{split}
\end{equation}
where the first two terms depend on $S$, and therefore we need to
compute the corresponding integral, while for the rest, the integral
is proportional to the PDF of the matrix--$t$ distribution, Eq.~\ref{eq:matrixtpdf}.
In order to compute:
\begin{equation}
n_{t}\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\log\det SdS\label{eq:firstlogpart}
\end{equation}
we note that taking the derivative with respect to $x$ in Eq.~\ref{eq:intmultgammadef},
we have:
\begin{equation}
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}\log\det SdS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}(\log\det A+\gamma_{p}(\frac{1}{2}(1+p+2x)))\label{eq:intderivative}
\end{equation}
We can solve Eq.~\ref{eq:firstlogpart} by substituting this result directly.

In order to compute the second nontrivial log term,
\begin{equation}
\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))dS
\end{equation}
we note that, due to the linearity of the trace, the result is equivalent
to solving a matrix valued integral
\begin{equation}
\textrm{tr}\left[\left(\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)SdS\right)(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\,.
\end{equation}
The enclosed integral also follows from Eq.~\ref{eq:intmultgammadef},
after taking the matrix derivative with respect to $A$. Using, for
symmetric $S$
\begin{equation}
\frac{d}{dA}\det A=(\det A)A^{-1}\label{eq:detderivative}
\end{equation}
and
\begin{equation}
\frac{d}{dA}e^{\textrm{tr}SA}=(e^{\textrm{trS}A})S
\end{equation}
we have 
\begin{equation}
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}SdS=-\frac{1}{2}(1+p+2x)\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(p+2x-1)}A^{-1}
\end{equation}
from where again, we can obtain the result by direct substitution.

@@

For $M_{t}$ we have:
\begin{equation}
\frac{dE}{dM_{t}}=0\Rightarrow\sum_{i}^{N}\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}\left(T_{i}-M_{t}\right)\Omega_{t}^{-1}=0
\end{equation}
and because $\Omega_{t}$ is nonsingular, we can obtain $M_{t}$ as
a weighted average of the input data:
\begin{equation}
M_{t}=\frac{1}{N}\sum_{i}^{N}\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}T_{i}=\frac{1}{N}\sum_{i}^{N}W_{i}T_{i}
\end{equation}
where we have defined the weight matrices $W_{i}$ as
\begin{equation}
W_{i}=\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}
\end{equation}

$\Sigma_{t}$ is given in terms of the inverse of the mean weight
by:
\begin{equation}
\Sigma_{t}=\frac{n_{t}+p-1}{n+m+p-1}\left[\frac{1}{N}\sum_{i}^{N}W_{i}\right]^{-1}
\end{equation}
Note that it depends on $n_{t}$.

For $\Omega_{t}$ we obtain:
\begin{equation}
\Omega_{t}=\frac{n+p+m-1}{p}\frac{1}{N}\sum_{i}^{N}(T_{i}-M_{t})W_{i}(T_{i}-M_{t})
\end{equation}
and for $n_{t}$, in the case where $\Sigma_{t}$ is a free parameter
that needs to be estimated, we have:
\begin{equation}
-p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log\det W_{i}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\end{equation}
otherwise if $\Sigma_{t}=\Sigma$ is fixed and does not depend on
$n_{t}$:
\begin{equation}
\log\det\Sigma_{t}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\end{equation}
